---
title: "Create data dictionary and Exploratory Data Analysis (EDA) for the HIV Dataset"
author: "Valerie Huanle Lin"
output: html_document
---


```{r}
#4.load the R packages required in the document and to set defaults for your document. 
library(dplyr)
library(dbplyr)
library(readr)
library(tidyverse)
library(Hmisc)
library(tsibbledata)
library(tsibble)
library(ggthemes)
library(gganimate)
library(gapminder)
library(ggplot2)
library(gganimate)
library(gridExtra)
library(rvest)
library(tibble)
library(janitor)
library(scales)
library(knitr)

```

```{r}
#5.read web url

url <- "https://figshare.com/ndownloader/files/35249488"
download.file(url, "hiv_dataset.csv")
data <- read.csv("hiv_dataset.csv")
str(data)
head(data)
summary(data)

```

```{r}
#6a) renaming columns so that the names do not include any spaces
data <- clean_names(data)
head(data)

#6b)converting categorical columns to factors, with appropriate labels for each factor value.
data$gender <- factor(data$gender, levels = 1:2, labels = c("Male", "Female"))
data$ethnic <- factor(data$ethnic, levels = 1:4, labels = c("Asian", "Afro", "Caucasian", "Other"))
data$base_drug_combo <- factor(data$base_drug_combo, levels = 0:5, labels = c("FTC + TDF", "3TC + ABC", "FTC + TAF", "DRV + FTC + TDF", "FTC + RTVB + TDF", "Other"))
data$comp_ini <- factor(data$comp_ini, levels = 0:3, labels = c("DTG", "RAL", "EVG", "Not Applied"))
data$comp_nnrti <- factor(data$comp_nnrti, levels = 0:3, labels = c("NVP", "EFV", "RPV", "Not Applied"))
data$extra_pi <- factor(data$extra_pi, levels = 0:5, labels = c("DRV", "RTVB", "LPV", "RTV", "ATV", "Not Applied"))
data$extra_pk_en <- factor(data$extra_pk_en, levels = 0:1, labels = c("False", "True"))
data$vl_m <- factor(data$vl_m, levels = 0:1, labels = c("False", "True"))
data$cd4_m <- factor(data$cd4_m, levels = 0:1, labels = c("False", "True"))
data$drug_m <- factor(data$drug_m, levels = 0:1, labels = c("False", "True"))

# View the converted data
glimpse(data)
```

#7.create data dictionary and statement

url <- "https://healthgym.ai/antiviral-hiv/"

# Read the webpage and extract the table
page <- read_html(url)
table_nodes <- page %>% html_nodes("table")  
table <- html_table(table_nodes[[1]], fill = TRUE, header = TRUE)  

# Convert the table to a data frame
df <- as.data.frame(table[, -1])


# Add a new column "description" to the data dictionary 
df$description <- c("Viral load measurement",
                  "Absolute CD4 count measurement",
                  "Relative CD4 count measurement",
                  "Gender of the patient",
                  "Ethnicity of the patient",
                  "Base drug combination used for treatment",
                  "Initial component of the treatment regimen",
                  "Non-nucleoside reverse transcriptase inhibitor component of the treatment regimen",
                  "Whether an extra protease inhibitor was included in the treatment regimen (True/False)",
                  "Whether an extra pharmacokinetic enhancer was included in the treatment regimen (True/False)",
                  "Whether viral load measurement was taken (True/False)",
                  "Whether CD4 count measurement was taken (True/False)",
                  "Whether drug measurement was taken (True/False)"
  )


# Display the data dictionary as a Markdown table
cat("Data Dictionary:\n")
knitr::kable(df, format = "markdown")


# Get the number of rows
num_rows <- nrow(table)

# Print the data dictionary
cat("\nNumber of Rows is:", num_rows, "\n")

```{r}
# Check for zero or negative values
any(data$vl <= 0, data$cd4 <= 0, data$rel_cd4 <=0)  # Check if any values are zero or negative

# Log-transform the variable
data$log_vl <- log(data$vl)
data$log_cd4 <- log(data$cd4)
data$log_rel_cd4 <- log(data$rel_cd4)

# Check the transformed variable
head(data[, c("log_vl", "log_cd4", "log_rel_cd4")])

# Histogram for log_vl
ggplot(data, aes(x = log_vl)) +
  geom_histogram(fill = "lightblue", color = "black", bins = 30) +
  labs(x = "Log VL", y = "Count", title = "Histogram of Log VL") +
  theme_minimal()
```

```{r}

# Histogram for log_cd4
ggplot(data, aes(x = log_cd4)) +
  geom_histogram(fill = "lightblue", color = "black", bins = 30) +
  labs(x = "Log CD4", y = "Count", title = "Histogram of Log CD4") +
  theme_minimal()

```

```{r}
# Histogram for log_rel_cd4
ggplot(data, aes(x = log_rel_cd4)) +
  geom_histogram(fill = "lightblue", color = "black", bins = 30) +
  labs(x = "Log Rel CD4", y = "Count", title = "Histogram of Log Rel CD4") +
  theme_minimal()
```


```{r}
# Plot the bar chart with gender counts
  ggplot(data, aes(x = gender)) +
  geom_bar(fill = "lightblue") +
  labs(x = "Gender", y = "Count", title = "Distribution of Gender") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  geom_text(stat = "count", aes(label = ..count..), vjust = -0.5)
```

```{r}
# Bar chart for Ethnic with counts
ggplot(data, aes(x = ethnic)) +
  geom_bar(fill = "lightgreen") +
  labs(x = "Ethnicity", y = "Count", title = "Distribution of Ethnicity") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  geom_text(stat = "count", aes(label = ..count..), vjust = -0.5)
```

```{r}
# Bar chart for Base Drug Combo with counts
ggplot(data, aes(x = base_drug_combo)) +
  geom_bar(fill = "lightyellow") +
  labs(x = "Base Drug Combo", y = "Count", title = "Distribution of Base Drug Combo") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  geom_text(stat = "count", aes(label = ..count..), vjust = -0.5)
```

```{r}
# Bar chart for Comp. INI with counts
ggplot(data, aes(x = comp_ini)) +
  geom_bar(fill = "lightpink") +
  labs(x = "Comp. INI", y = "Count", title = "Distribution of Comp. INI") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  geom_text(stat = "count", aes(label = ..count..), vjust = -0.5)
```

```{r}
# Bar chart for Comp. NNRTI with counts
ggplot(data, aes(x = comp_nnrti)) +
  geom_bar(fill = "orange") +
  labs(x = "Comp. NNRTI", y = "Count", title = "Distribution of Comp. NNRTI") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  geom_text(stat = "count", aes(label = ..count..), vjust = -0.5)
```

```{r}
# Bar chart for Extra pk-En with counts
ggplot(data, aes(x = extra_pk_en)) +
  geom_bar(fill = "lightgray") +
  labs(x = "Extra pk-En", y = "Count", title = "Distribution of Extra pk-En") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  geom_text(stat = "count", aes(label = ..count..), vjust = -0.5)
```


```{r}
# Bar chart for VL (M) with counts
ggplot(data, aes(x = vl_m)) +
  geom_bar(fill = "lightcyan") +
  labs(x = "VL (M)", y = "Count", title = "Distribution of VL (M)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  geom_text(stat = "count", aes(label = ..count..), vjust = -0.5)
```

```{r}
# Bar chart for CD4(M) with counts
ggplot(data, aes(x = cd4_m)) +
  geom_bar(fill = "lightblue") +
  labs(x = "CD4(M)", y = "Count", title = "Distribution of CD4(M)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  geom_text(stat = "count", aes(label = ..count..), vjust = -0.5)
```


```{r}
# Bar chart for Drug Recorded (Drug (M)) with counts
ggplot(data, aes(x = drug_m)) +
  geom_bar(fill = "lightblue") +
  labs(x = "Drug Recorded (Drug (M))", y = "Count", title = "Distribution of Drug Recorded (Drug (M))") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  geom_text(stat = "count", aes(label = ..count..), vjust = -0.5)
```

```{r}
#line plot for timepoints
ggplot(data, aes(x = timepoints)) +
  geom_freqpoly() +
  labs(x = "Timepoints", y = "Count") +
  ggtitle("Timepoints Diagram")

```
 
 
```{r}  
# Grouped Bar Chart: Base Drug Combo by Gender 
ggplot(data, aes(x = factor(gender), fill = factor(base_drug_combo))) +
  geom_bar(position = "fill") +
  labs(x = "Gender", y = "Proportion", fill = "Base Drug Combo") +
  theme_minimal()
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) 
  geom_text(aes(label = scales::percent(..count../sum(..count..))), 
            position = position_fill(vjust = 0.5))
```

```{r}  

#Relationship between VL and CD4
ggplot(data, aes(x = vl, y = cd4)) +
  geom_point(color = "steelblue", size = 3, alpha = 0.7) +
  labs(x = "VL (Viral Load)", y = "CD4 Count") +
  ggtitle("Relationship between VL and CD4") +
  theme_minimal()


```


